---
title: "Detection thresholds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{detection_thresholds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(foldchanger)
library(dplyr)
library(ggplot2)
```

Analytical pipelines require detection thresholds for gene dosage. The application
of thresholds allows the pipeline to classify a gene region as a particular
dosage state, such as "normal", "gain" or "loss".

However, determining the sensitivity of a threshold can be challenging. This is
because the accuracy of the threshold will depend on the sample's neoplastic cell 
content (NCC) and the number of gene copies per cell; both of these values vary
between samples.

Using the `pred_cpc` function, we can plot detection limit for a fold change 
threshold of 2.8.

```{r, echo = FALSE}

df <- data.frame(ncc = seq(0, 100, by = 0.1)) |> 
  mutate(gene_copies = pred_cpc(fc = 2.8, ncc = ncc, dir = "gain",
                                digits = 3))

gain_lod_plot <- df |> 
  filter(gene_copies <= 100) |> 
  ggplot(aes(x = gene_copies, y = ncc)) +
  geom_line(linewidth = 1) +
  theme_bw() +
  labs(x = "Gene copies per cell", y = "Neoplastic cell content (%)",
       title = "Detection limit for gains",
       subtitle = "Fold change threshold of 2.8") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0,100,10),
                     minor_breaks = FALSE) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0,100,10),
                     minor_breaks = FALSE) +
  annotate(geom = "text", x = 30, y = 30, label = "Detected") +
  annotate(geom = "text", x = 30, y = 5, label = "Not detected") 

gain_lod_plot

```


