---
title: "Detection thresholds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{detection_thresholds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(foldchanger)
library(dplyr)
library(ggplot2)
library(tidyr)
```

Analytical pipelines require detection thresholds for gene dosage. The application
of thresholds allows the pipeline to classify a gene region as a particular
dosage state, such as "normal", "gain" or "loss".

However, determining the sensitivity of a threshold can be challenging. This is
because the accuracy of the threshold will depend on the sample's neoplastic cell 
content (NCC) and the number of gene copies per cell; both of these values vary
between samples.

# Amplifications

In cancer, oncogenes can be amplified to copy numbers rangeing from tens to
hundreds of copies above the standard 2 copies per cell. Using the `pred_cpc` 
function, we can predict the gene copies per tumour cell
which would be detected at different neoplastic cell contents when a fold
change threshold of 2.8 is applied.

```{r, include = TRUE}

df_amp <- data.frame(ncc = seq(0, 100, by = 0.1)) |> 
  mutate(gene_copies = pred_cpc(fc = 2.8, 
                                ncc = ncc, 
                                dir = "gain",
                                digits = 3))

```

We can then use `ggplot` to show the results.

```{r, echo = FALSE}
gain_lod_plot <- df_amp |> 
  filter(gene_copies <= 100) |> 
  ggplot(aes(x = gene_copies, y = ncc)) +
  geom_line(linewidth = 1) +
  theme_bw() +
  labs(x = "Gene copies per cell", y = "Neoplastic cell content (%)",
       title = "Detection limit for gene amplifications",
       subtitle = "Fold change threshold of 2.8") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0,100,10),
                     minor_breaks = FALSE) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0,100,10),
                     minor_breaks = FALSE) +
  annotate(geom = "text", x = 30, y = 30, label = "Detected") +
  annotate(geom = "text", x = 30, y = 5, label = "Not detected") 

gain_lod_plot
```

# Deletions

For gene deletions, the number of gene copies per tumour cell is more limited.
Assuming that the cell is diploid, the cell can have a normal gene copy number
of 2, or a copy number of 1 if the gene has been heterozygously deleted, or a 
copy number of 0 if the gene has been homozygously deleted.

By using the `pred_fc` function, we can predict the fold changes for each
deletion state at different neoplastic cell contents.

```{r, include = TRUE}
df_del <- data.frame(ncc = seq(0, 100, by = 0.1)) |> 
  mutate(fold_change_normal = pred_fc(tumour_cpc = 2,
                                      ncc = ncc,
                                      dir = "loss",
                                      digits = 3),
         fold_change_het_del = pred_fc(tumour_cpc = 1,
                                       ncc = ncc,
                                       dir = "loss",
                                       digits = 3),
         fold_change_hom_del = pred_fc(tumour_cpc = 0,
                                       ncc = ncc,
                                       dir = "loss",
                                       digits = 3)) |> 
  pivot_longer(cols = -ncc,
               names_to = "del_state",
               values_to = "fold_change")
```

By plotting the results, we can see that determination of the neoplastic cell
content is critical for differentiating between homozygous and heterozygous
deletions when the fold change is greater than -2.

```{r, echo = FALSE}
del_fc_plot <- ggplot(df_del, aes(x = ncc, y = fold_change)) +
  geom_line(aes(colour = del_state), linewidth = 1) +
  scale_colour_manual(values = c( "#999999", "#999999", "#999999")) +
  coord_cartesian(ylim = c(-5, 0)) +
  scale_y_continuous(breaks = c(0, -1, -2, -3, -4, -5),
                     minor_breaks = NULL) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  theme_bw() +
  theme(legend.position = "blank") +
  labs(y = "Fold change (loss equation)", x = "Neoplastic cell content",
       colour = "") +
  annotate("text", x = 65, y = -4.5, label = "Homozyous \n deletion") +
  annotate("text", x = 90, y = -2.5, label = "Heterozygous \n deletion") +
  annotate("text", x = 90, y = -0.5, label = "Normal")

del_fc_plot
```
